---
title: "Take-home Exercise 1"
author: "Chrissandro"
date: "14 January 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
editor: visual
---

# 1. Project Overview

## 1.1. Setting the Scene

OECD education director Andreas Schleicher shared that *"Singapore managed to achieve excellence without wide differences between children from wealthy and disadvantaged families"*. Singapore's Minister for Education shares the same belief, starting an *"every school a good school"* slogan. However, the population of Singapore disagrees that disparities don't exist, especially **between elite and neighborhood schools; students from higher socioeconomic status families and lower ones; and immigration and non-immigration families**.

## 1.2. The Task

Based on the dataset provided by the 2022 Programme for International Student Assessment (PISA), the aim of this exercise is to utilize appropriate **Exploratory Data Analysis (EDA)** methods and **ggplot2** functions to reveal:

-   the distribution of Singapore students' performance in mathematics, reading, and science, and

-   the relationship between these performances with schools, gender and socioeconomic status of the students.

# 2. Data Preparation

## 2.1 Installing R packages

In this take-home exercise, two R packages will be used. They are:

-   [tidyverse](https://www.tidyverse.org/), and

-   [haven](https://www.haven.tidyverse.org/)

The code chunk used is as follow:

```{r}
pacman::p_load(tidyverse, haven)
```

## 2.2 Importing PISA data

The code chunk below uses [`read_sas()`](https://www.haven.tidyverse.org/reference/read_sas.html) of [haven](https://www.haven.tidyverse.org/) to import PISA data into R environment.

```{r}
stu_qqq <- read_sas("data/cy08msp_stu_qqq.sas7bdat")
```

To retrieve only relevant data (Singapore-based), the *filter* argument is used as seen in the code chunk below.

```{r}
stu_qqq_SG <- stu_qqq %>%
  filter(CNT == "SGP")
```

Then, the *write_rds* function will generate a new .rds file to be used for consequent EDAs.

```{r}
write_rds(stu_qqq_SG,
          "data/stu_qqq_SG.rds")
```

```{r}
stu_qqq_SG <-
  read_rds("data/stu_qqq_SG.rds")
```